Type Classes for Mathematics in Type Theory

### 2010-12-14 papers/mscs/mscs.tex
Fitting into 30pp

% \documentclass{mscs}
\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{harvard}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}

\usepackage{color}

@@

Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident. It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable procedures and data structures, implemented using type theory's native term reduction as a programming language. Thus, our work very much belongs in the long tradition of trying to realize the promise of type theory to truly unite mathematical formalization and certified (functional) programming, without making painful sacrifices on either side.

@@

\paragraph{Outline.}
In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes. Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\newpage
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: ∀ x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

In the next section, we show that type-classification of structure predicates and their component parameters has the potential to remedy the problems associated with the naive unbundled predicate approach, giving us the best of both worlds.

@@

Next, we turn to problems concerning theorem \emph{declaration}. As a reference point, our ideal will be the common mathematical vernacular, where one simply says:
\begin{quote}
Theorem: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
\end{quote}
(This silly statement allows us to clearly present the syntax.)

@@

While we are on the topic of implicit generalization, we mention one inadequacy concerning their current implementation that we feel should be addressed for the facility to be a completely satisfying solution. While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise". We have found that the need for such a policy arises naturally when declaring a parameter of class type in a context where \emph{some} of its components are already available, while others are to be newly introduced. The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.

@@

One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context. For the example theorem above, the context looks like:
\begin{lstlisting}
  G: Type
  e: Equiv G
  op: SemiGroupOp G
  P: SemiGroup G e op
\end{lstlisting}
We are not particularly worried about overly large contexts, especially because most of the `extra' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense inert with respect to proof search. Hence, we do not foresee problems with large contexts for any but the most complex formalizations.

@@

We have developed an algebraic hierarchy composed entirely out of predicate classes and operational classes as described in the previous section. For instance, our semiring interface looks as follows:
\begin{lstlisting}
  Class SemiRing A {e: Equiv A}
      {plus: RingPlus A} {mult: RingMult A}
      {zero: RingZero A} {one: RingOne A}: Prop :=
    { semiring_mult_monoid:> CommutativeMonoid A (op:=mult)(unit:=one)
    ; semiring_plus_monoid:> CommutativeMonoid A (op:=plus)(unit:=zero)
    ; semiring_distr:> Distribute mult plus
    ; semiring_left_absorb:> LeftAbsorb mult zero }.
\end{lstlisting}
All of \lstinline|Equiv|, \lstinline|RingPlus|, \lstinline|RingMult|, \lstinline|RingZero|, and \lstinline|RingOne| are operational (single-field) classes, with bound notations \lstinline|=|, \lstinline|+|, \lstinline|*|, \lstinline|0|, and \lstinline|1|, respectively.

Let us briefly highlight some additional aspects of this style of structure definition in more detail.

Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive). In our hierarchy, these substructures by themselves establish the following inheritance diagram:

\vspace{5mm}
\begin{center}
\includegraphics{hierarchy.pdf}
\end{center}
\vspace{5mm}

@@

\paragraph{Acknowledgements.}
This research was stimulated by the new possibilities provided by the introduction of type classes
in Coq. In fact, this seems to be the first substantial development that tries to exploit all their
possibilities. As a consequence, we found many small bugs and unintended behavior in the type
class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
Georges Gonthier and Claudio Sacerdoti Coen have helped to
sharpen our understanding of the relation with Canonical Structures and with Unification Hints. Finally, James McKinna suggested many corrections and improvements.

This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
% MSCS recommends harvard, but does not specify which bst file to use.
\bibliographystyle{agsm}
\bibliography{alg}
\end{document}

>>>>

% \documentclass{mscs}
\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
% \usepackage{harvard}
\usepackage{amssymb}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
\usepackage{wrapfig}
\usepackage{color}

@@

Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident. It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable procedures and data structures, implemented using type theory's native term reduction as a programming language. Thus, our work belongs in the long tradition of realize the promise of type theory to truly unite mathematical formalization and certified (functional) programming, without making painful sacrifices on either side.

@@

\emph{Outline.}
In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes. Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:

\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: ∀ x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

In the next section, we show that type-classification of structure predicates and their component parameters has the potential to remedy the problems associated with the naive unbundled predicate approach.

@@

Next, we turn to problems concerning theorem \emph{declaration}. Our ideal will be the common mathematical vernacular, where one simply says:
\begin{quote}
Theorem: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
\end{quote}
(This silly statement allows us to clearly present the syntax.)

@@

While we are on the topic of implicit generalization, we mention one inadequacy concerning their current implementation that we feel should be addressed for the facility to be a completely satisfying solution. While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise". The need for such a policy arises naturally when declaring a parameter of class type in a context where \emph{some} of its components are already available, while others are to be newly introduced. The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.

@@

One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context. For the theorem above, the context looks like:
\begin{lstlisting}
  G: Type
  e: Equiv G
  op: SemiGroupOp G
  P: SemiGroup G e op
\end{lstlisting}
We are not particularly worried about overly large contexts, especially because most of the `extra' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense inert with respect to proof search. Hence, we do not foresee problems with large contexts for any but the most complex formalizations.

@@

We have developed an algebraic hierarchy composed entirely out of predicate classes and operational classes as described in the previous section. For instance, our semiring interface looks as follows:
\begin{lstlisting}
  Class SemiRing A {e: Equiv A}
      {plus: RingPlus A} {mult: RingMult A}
      {zero: RingZero A} {one: RingOne A}: Prop :=
    { semiring_mult_monoid:> CommutativeMonoid A (op:=mult)(unit:=one)
    ; semiring_plus_monoid:> CommutativeMonoid A (op:=plus)(unit:=zero)
    ; semiring_distr:> Distribute mult plus
    ; semiring_left_absorb:> LeftAbsorb mult zero }.
\end{lstlisting}

All of \lstinline|Equiv|, \lstinline|RingPlus|, \lstinline|RingMult|, \lstinline|RingZero|, and \lstinline|RingOne| are operational (single-field) classes, with bound notations \lstinline|=|, \lstinline|+|, \lstinline|*|, \lstinline|0|, and \lstinline|1|, respectively.
\begin{wrapfigure}{r}{.4\textwidth}
    \includegraphics[scale=.8]{hierarchy.pdf}
%   \caption{}
\end{wrapfigure}
Let us briefly highlight some additional aspects of this style of structure definition in more detail.

Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive). In our hierarchy, these substructures by themselves establish the inheritance diagram.

@@

\emph{Acknowledgements.}
This research was stimulated by the new possibilities provided by the introduction of type classes
in Coq. In fact, this seems to be the first substantial development that tries to exploit all their
possibilities. As a consequence, we found many small bugs and unintended behavior in the type
class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
Georges Gonthier and Claudio Sacerdoti Coen have helped to
sharpen our understanding of the relation with Canonical Structures and with Unification Hints. Finally, James McKinna suggested many improvements.

This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
% MSCS recommends harvard, but does not specify which bst file to use.
\bibliographystyle{plain}
\bibliography{alg}
\end{document}
